Epistemic Semantics in Guarded String Models

### 2020-09-19 tex/sdraft.tex
Cuts in sdraft

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage{times}
\usepackage{latexsym}
\usepackage{amssymb,amsmath,accents}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{hang}
\renewcommand{\UrlFont}{\ttfamily\small}
\usepackage{microtype}

@@

\pagestyle{plain}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.

% Uncomment this line for the final submission
% Or comment it for the submission version with line numbers.
\aclfinalcopy 
%\def\aclpaperid{***} %  Enter the acl Paper ID here
% Make the references come out black. TAKE THIS OUT for submission.
\definecolor{darkblue}{rgb}{0, 0, 0}
% CHECK THE BIB STYLE FOR THE STARSEM SUBMISSION

% This used to say Constructive ..., which is a better title.
% Reserve that for a journal paper if we get to that.
\title{Epistemic Semantics in Guarded String Models}

\author{Eric Campbell \\
  Cornell University\\
  \texttt{ehc86@cornell.edu} \\\And
  Mats Rooth \\
  Cornell University\\
  \texttt{mr249@cornell.edu} \\}

@@

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}. The models we propose are an
epistemic extension of guarded string models for KAT, where a unary
operation interpreted as an existential epistemic modality is included
for each agent.  {\em Action models} in dynamic epistemic semantics
introduce the technique of constructing epistemic models from
primitive alternative relations on events, in order to capture the
epistemic consequences of perceptual and communicative events
\citep{baltag1999logic}. This is the basis for our construction of
epistemic alternative relations.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensiononal semantics} has investigated how to do the
semantics of intensional complementation, including indirect
questions, in a setting where compositional semantics is expressed in
a finite state calculus \citep{rooth2017finite,collard2018finite}.  We
adopt this in our syntax-semantics interface for English.

@@

\begin{example} \label{coin2}
\hspace{-4mm} \begin{tabular}[t]{ll}
   $w_1$ \\
   $w_2 = w_1 \hspace{2pt} a_1 $ \\
   $w_2 = w_1 \hspace{2pt} a_1 \hspace{2pt} b_1 $\\ 
 \end{tabular}
\end{example}

@@

The events come with pre-conditions.
Amy can peek at heads only if the coin is heads up, so $a_1$ has the precondition of
the coin being heads up.  Let $h$ be the Boolean proposition that the
coin is heads up.
Then preconditions
can be described by Boolean formulas, with $h$ being the precondition
of $a_1$
$u$. Events come as well with a relation between
prior and following state, for instance with $u$ incrementing the
floor.  This is expressed using an operator ``$:$'' (read ``and
next'') that pairs Boolean formulas. The first line in \ex{event1}
describes $a_1$ (Amy looking at heads) as happening only in an $h$
state, and as not changing the state.  Symmetrically, $a_0$ (Amy
looking at tails) can happen only in a not-$h$ state, and does not
change the state.

@@

\renewcommand{\arraystretch}{1.0}

\begin{example}   \label{event1}
\hspace{-5mm}\( \begin{array}[t]{cl}
a_1 & h:h \\
a_0 & \ev{\bar{h}}{\bar{h}} \\
% u & (\bar{q}\bar{p}) \! : \!(\bar{q}p) + \\
% u & \ev{(\bar{q}\bar{p})}{(\bar{q}p)} + \ev{(\bar{q}p)}{(q\bar{p})} + \ev{(q\bar{p})}{(qp)} \\
% d & \evr{(\bar{q}\bar{p})}{(\bar{q}p)} + \evr{(\bar{q}p)}{(q\bar{p})} + \evr{(q\bar{p})}{(qp)} \\ 
\end{array} \)
\end{example}

@@

Well-formed guarded strings have the interpretation of worlds in the
application to natural-language semantics.  The set of possible worlds
in the Kripke frame determined by an Epik specification is the set of
well-formed guarded strings.  At this point, we could say that any set
of worlds is a proposition, so that the set of propositions is the
power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead
define a more restrictive set of propositions corresponding to the
regular sets of strings. This is deferred to the next section.

In the natural language application, sets of well-formed guarded strings 
have the addional interpretation of event types.
An event-type is something that can ``happen'' is different worlds.
The event of the elevator going up is modeled not as the bare event
symbol $u$ or its unit set $\{u\}$, but as the set of guarded strings
\(\{ \hat 0 \,u\, \hat 1, \hat 1 \,u\, \hat 2,  \hat 2 \,u\, \hat 3 \}\).
The event of the elevator going up in a given world $w$ corresponds
to $w$ being incremented to form $w \diamond e$, where $e$ is an element
of the event type.  \ex{events1} gives the event types in
the examples.

\begin{example} \label{events1}
  \renewcommand{\arraystretch}{1.1}
\(
\begin{array}[t]{l|c}
   &  {\mbox{Event type}}\\ \hline
  u & \{ \hat 0 \,u\, \hat 1, \hat 1 \,u\, \hat 2,  \hat 2 \,u\, \hat 3\}\\
  d & \{ \hat 1 \,d\, \hat 0, \hat 2\,d\,\hat 1, \hat 3 \,d\, \hat 2\}\\
  a_1 & \{H \,a_1\, H\}\\
  a_0 & \{T \,a_0\, T\}\\
  b_1 & \{H \,b_1\, H \}\\
  b_0 & \{T \,b_0\, T \}
\end{array}
\)
  \renewcommand{\arraystretch}{1}
\end{example}

The construction so far defines a set of worlds from an Epik specification.  Normally
and in our examples, $W$ is countably infinite, though some choices of effect
formulas can result in a finite set of worlds.  The next step is to define an
alternative relation $R_a$ on worlds for each agent $a$. This will result in a Kripke
frame $\tuple{W,R_1,...,R_n}$ consisting of a set of worlds, and a world-alternative
relation for each agent \citep{kripke1963semantical}.  An Epik specification defines
an alternative relation on bare events for each agent $a$, which we notate as
${R}_{a}$.  This should be lifted to a relation $\hat{R}_a$ on worlds.  The basic
idea is that when a world $w$ is incremented with an event $e$, in the resulting
world $w \diamond e$, epistemic alternatives for agent $a$ are of the form $w'
\diamond e'$, where $w'$ is an alternative to for $a$ in $w$, and $e'$ is and
event-alternative to $e$ for $a$.\footnote{In this it is important that the
  event-alternative relation for an agent is constant across worlds. We anticipate
  that the definition given here produces results equivalent to what is found in
  literature on event alternatives in dynamic epistemic semantics, though we have not
  verified this.  That literature primarily focuses on mapping an epistemic model for
  a single time and situation to another, and uses general first-order models, rather
  than guarded string models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and
  articles in \citet{ditmarsch2015handbook}.  We picked up the idea from papers by
  Moss and his colleagues, together with \citet{moore1984formal} and subsequent
  literature in situation theory.  This previous literature is motivated by epistemic
  logic and AI planning, rather than natural language semantics.  } This needs to be
implemented in a way that takes account of pre- and post-conditions for events. For
this, our approach is to refer the definition of well-formed guarded strings.
\ex{raise1} defines a relation on worlds from a relation on bare events.

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over  events P and primitive tests B, and ${R}$ be a relation
  on $P$. The corresponding relation $\hat{R}$ on $W$ holds between a guarded string
\[
\alpha_0e_0 ... e_n\alpha_{n+1} 
\]
in $W$ and a guarded string  $q$ iff $q$ is an element of $W$ and is of the form

\[
\alpha_0'e_0' ... e_n'\alpha_{n+1}', 
\]

where for $0 \leq n$, $\element{\tuple{e_i,e'_i}}{\hat{R}}$.
\end{example}

This requires that in the alternative world $q$, each constituent event $e'_i$ is an
alternative to the event $e_i$ in the base world according to $\hat{R}$.
Compatibilities between events in the alternative world are enforced by the
requirement that the alternative $q$ is an element of $W$, so that state and effect
formulas are enforced.

@@

\section{The logical language of Epistemic KAT}
In addition to a set of possible worlds and an algebra of propositions, the
application in natural language requires a logical language for naming propositions.
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for KAT, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  }. To this signature we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  The intended interpretation of $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used in place of the product symbol $\cdot$.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet *$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet \! * \! h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet\!*\!h$ is the set
of worlds where the coin is heads.\footnote{A mapping from English sentences to logical
  terms in epistemic KAT is presented in Section 6}.

@@

@@

Each term in the language is interpreted as a set of guarded strings.  Where $p$ is a
term, $\sem{p}^{\epikspec}$ is the interpretation of the term, where superscript
captures dependence on an Epik specification.  Figure \ref{fig:interp2} defines the
interpretation of terms recursively.  The interpretation $\sem{ 1 }^{\epikspec}$ of
the multiplicative identity is the set of atoms that satisfy the state constraint
$\varphi$.  Where $b$ is a primitive Boolean, $\sem{ b }^{\epikspec}$ is the set of
atoms that satisfy the state constraint and where $b$ is true.  Where $e$ is a
primitive event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have the
form of $e$ flanked by compatible atoms, as determined by the event formula
$\zeta_e$. The product $pq$ of terms $p$ and $q$ is interpreted with fusion product
raised to sets of guarded strings.
Kleene star is interpreted as the union of iterated products. $p^n$ is the $n$-times product
of $p$ with itself, with $p^0 = 1$.
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of the
complement proposition under the relation
$\hat{R}_a$.  Here $\hat{R}_a$ is the world-alternative for agent $a$,
as defined in \ex{raise1}.

@@

Summing up, given an Epik specification, a state constraint $\varphi$, and effect
relation $\zeta_e$ for each primitive are constructed. Given these, each term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$ over the Boolean primitives and event primitives in the
specification.  Let $K^{\epikspec}$ be the sets that are interpretations of terms.
Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of epistemic KAT, with
operations as in Figure \ref{fig:interp2} (e.g. the binary operation $+$ is union,
and the unary operation $\Diamond_a$ is pre-image relative to $\hat{R}_a$).  The
construction can also be viewed as providing a concrete $n$-agent Kripke frame
$\tuple{\sem{\bullet *}^{\epikspec}, \hat{R}_1,...,\hat{R}_n}$.\footnote{ The domain
  of the Kripke frame differs from the domain of the guarded string model, because
  the former is the set of worlds, while the latter is the set of propositions.  }
The frame consists of a set of worlds, and an epistemic-alternative relation for each
agent.  These models are used as a target for natural-language interpretation in
Section 5 and Section 6.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of some of
the worlds from the examples.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
multi-character symbols that are found in implementations of the finite state
calculus. The multi-character symbol {\tt a0} is an element of the alphabet that has
no connection with the element {\tt a}.

 \hspace{-13mm} \begin{tabular}[t]{cl}
{\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
$v_1$ & {\tt 0 0 u 0 1 u 1 0 d  0 1 u 1 0} \\
$v_2$ & {\tt 0 0 u 0 1 u 1 0 u 1 1} \\
$w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 10} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

Terms in the finite state calculus are interpreted as sets of strings, or for
relational terms, as relations between strings.  Computationally, the sets and
relations are represented by finite state acceptors.  As used here, a program in the
Fst language of the finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and functions mapping
one or more regular relations to a regular set or relation.  The functions are
defined as macros.  To illustrate, \ex{stex1} is part of an Epik program that
declares states and a state constraint.  This is equivalent to the Fst program
\ex{stex2}. The result of the sequence of definitions is that {\tt St} is the set of
Boolean vectors or length four that are well-formed according to the state formula;
Nst is the complement operation for tests; H1 is the set of well-formed Boolean
vectors that have 1 in the first position; and so forth.  Translation from an Epik
program to an Fst program is accomplished in Haskell.

\begin{example}\label{stex1}
Part of an Epik program
\end{example}

\vspace{-3mm} {\small
\begin{verbatim}  
   state H1 T1 H2 T2
   constraint !(H1 T1) & (H1 + T1) 
              !(H2 T2) & (H2 + T2) 
\end{verbatim}}

\begin{example}\label{stex2}
Corresponding Fst program
\end{example}

\vspace{-3mm}
{\small 
\begin{verbatim} 
   define Bool ["0"|"1"];
   define St0 [Bool Bool Bool Bool];
   define H1 [1 Bool Bool Bool];
   define T1 [Bool 1 Bool Bool];
   define H2 [Bool Bool 1 Bool];
   define T2 [Bool Bool Bool 1];
   define St St0 & ((((St0 - (H1 & T1)) 
                 & (H1 | T1)) 
                 & (St0 - (H2 & T2))) 
                 & (H2 | T2));
   define Nst(X) [St - X];
   define H1 St & H1;
   define T1 St & T1;
   define H2 St & H2;
   define T2 St & T2;
\end{verbatim}}

Event symbols are interpreted as event types, in the way discussed in connection with
\ref{events1}.  For instance, defines an event type {\tt paT2type} of Amy peeking at the second coin
in a scenario with two coins, and seeing that it is tails.\footnote{After the definition,
  {\tt paT2type} is a constant symbol denoting a set of two strings. On the right hand side of the definition,
  {\tt paT2} is the unit set of the symbol {\tt paT2}.  
}  The first conjunct
expresses that this event can only happen if the second coin is tails ({\tt H2}), and that the
state of the second coin does not change.  The second conjunct expresses that the
event does not change the state of the first coin -- either it is heads ({\tt H1}) and
remains heads, or it is tails ({\tt T1}) and remains tails.  Such formulas are mapped from
effect formulas in Epik programs. The state formula is imposed as well, since it
figures in the definition of state constants such as {\tt H1}.

\begin{example} \label{peek1}
{\small 
\begin{verbatim}
define paT2type [[[T2 paT2] T2] & 
  [[[H1 paT2] H1] | [[T1 paT2] T1]]];
\end{verbatim} }
\end{example}

Semantically in Fst, {\tt paT2type} is a set of two strings, which can be displayed
in an interpreter for the language by reading {\tt paT2type} as a regular expression
and printing its extension, see \ex{printwords1}.

\begin{example}
\begin{verbatim}
xfst[1]: regex paT2type;
3.0 Kb. 16 states, 16 arcs, 2 paths.
xfst[2]: print words
0 1 0 1 paT2 0 1 0 1
1 0 0 1 paT2 1 0 0 1
\end{verbatim}
\end{example}

The finite state calculus has a product operation of string concatenation raised to sets.
Concatenation of strings with atoms (Boolean vectors) at both ends has the effect of
doubling atoms at the juncture, and does not enforce matching of atoms at the juncture.  Therefore
KAT product can not be identified with product in the finite state calculus. Instead,
KAT product and KAT Kleene star are defined operations, see Figure \ref{fstfigure1}.
The binary product operation {\tt Cn} and the unary Kleene star operation
{\tt Kst} combine strings in the string algebra, remove strings with
non-matching atoms, and then delete the second of two tests to create well-formed
guarded strings.  Matching of atoms is enforced with the set {\tt Wf0}, which is
the set of strings that do not contain non-matching Boolean vectors.  The containment
operator (expressed by the dollar sign) and complement (expressed by tilde) are operators of the finite
state calculus.  The set of non-matching sequences of atoms {\tt UnequalStPair} is defined by a finite disjunction.
Deletion is accomplished by a re-write rule in the finite state calculus, which is a
notation for defining regular relations by contextually constrained substitutions
(REFS).  In this case, is {\tt Squash} us a regular relation that deletes an atom
(a sequence of 0's and 1's of a certain length) when it follows an atom.\footnote{
  This is a non-equal length regular relation. The finite state calculus includes such
  relations, and they can be used with relation composition and relation domain and co-domain.
  They are restricted in that the complement and set difference for non-equal length relations is not defined.
The epistemic alterative relations that are defined in Figure -- are equal-length relations.
  }

@@

A given bare event such as $a_1$ (Aly looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, semantically $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$, and an algebra of
propositions as regular sets of strings drawn from this space of worlds.

\vspace{-3mm}
\begin{example} \label{alyrel1}
 \(
 \left\{
 \begin{array}{l}
   \pair{10a_110}{10a_110}, \\  \pair{01a_001}{01a_001}, \\
   \pair{10b_110}{10b_110}, \\ \pair{10b_110}{01b_001}, \\
   \pair{01b_001}{10b_110}, \\ \pair{01b_001}{01b_001} \\
 \end{array}
 \right\}
 \) 
\end{example}

@@

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage{times}
\usepackage{latexsym}
%\usepackage{caption}
\renewcommand{\UrlFont}{\ttfamily\small}

%\captionsetup{belowskip=0pt}

%\setlength{\abovecaptionskip}{-10pt}
\setlength{\belowcaptionskip}{-15pt}

\setlength{\abovecaptionskip}{5pt plus 3pt minus 2pt}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

%\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

\title{Epistemic Semantics in Guarded String Models}

%\author{Eric Campbell \\
%  Cornell University\\
%  \texttt{ehc86@cornell.edu} \\\And
%  Mats Rooth \\
%  Cornell University\\
%  \texttt{mr249@cornell.edu} \\}

\usepackage{amssymb,amsmath,accents}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{hang}
\renewcommand{\UrlFont}{\ttfamily\small}
\usepackage{microtype}

@@

@@

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.  {\em Action models} in dynamic
epistemic semantics introduce the technique of constructing epistemic
models from primitive alternative relations on events, in order to
capture the epistemic consequences of perceptual and communicative
events \citep{baltag1999logic}.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensiononal semantics} has investigated how to do the
semantics of intensional complementation in a setting where
compositional semantics is expressed in a finite state calculus
\citep{rooth2017finite,collard2018finite}.

@@

\begin{example} \label{coin2}
   $w_2 = w_1 \hspace{2pt} a_1 $ \hspace{2mm}
   $w_2 = w_1 \hspace{2pt} a_1 \hspace{2pt} b_1 $ 
\end{example}

@@

@@

The events come with pre-conditions.
Amy can peek at heads only if the coin is heads up, so $a_1$ has the precondition of
the coin being heads up.  Let $h$ be the Boolean proposition that the
coin is heads up.
Then preconditions
can be described by Boolean formulas, with $h$ being the precondition
of $a_1$
$u$. Events come as well with a relation between
prior and following state, for instance with $u$ incrementing the
floor.  This is expressed using an operator ``$:$'' (read ``and
next'') that pairs Boolean formulas. The formula $h:h$
describes $a_1$ (Amy looking at heads) as happening only in an $h$
state, and as not changing the state.  Symmetrically, $a_0$ (Amy
looking at tails) can happen only in a not-$h$ state, and does not
change the state ($\ev{\bar{h}}{\bar{h}}$).

%\renewcommand{\arraystretch}{1.0}

@@

Well-formed guarded strings have the interpretation of worlds in the
application to natural-language semantics.  The set of possible worlds
in the Kripke frame determined by an Epik specification is the set of
well-formed guarded strings.  At this point, we could say that any set
of worlds is a proposition, so that the set of propositions is the
power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead
define a more restrictive set of propositions corresponding to the
regular sets of strings. This is deferred to the next section.
Certai sets of well-formed guarded strings 
have the addional interpretation of event types.
An event-type is something that can ``happen'' is different worlds.
\ex{events1} gives the event types in
the example.

\begin{example} \label{events1}
  \renewcommand{\arraystretch}{1.1}
\(
\begin{array}[t]{l|cl|c}
  &  {\mbox{Event type}} && {\mbox{Event type}} \\ \hline
  a_1 & \{H \,a_1\, H\} &   b_1 & \{H \,b_1\, H \} \\
  a_0 & \{T \,a_0\, T\} &   b_0 & \{T \,b_0\, T \} \\
\end{array}
\)
  \renewcommand{\arraystretch}{1}
\end{example}

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect
formulas can result in a finite set of worlds.  The next step is to define an
alternative relation $R_a$ on worlds for each agent $a$. This will result in a Kripke
frame $\tuple{W,R_1,...,R_n}$ consisting of a set of worlds, and a world-alternative
relation for each agent \citep{kripke1963semantical}.  An Epik specification defines
an alternative relation on bare events for each agent $a$, which we notate as
$\hat{R}_{a}$.  This should be lifted to a relation ${R}_a$ on worlds.  The basic
idea is that when a world $w$ is incremented with an event $e$, in the resulting
world $w \diamond e$, epistemic alternatives for agent $a$ are of the form $w'
\diamond e'$, where $w'$ is an alternative to for $a$ in $w$, and $e'$ is and
event-alternative to $e$ for $a$.\footnote{In this it is important that the
  event-alternative relation for an agent is constant across worlds. We anticipate
  that the definition given here produces results equivalent to what is found in
  literature on event alternatives in dynamic epistemic semantics, though we have not
  verified this.  That literature primarily focuses on mapping an epistemic model for
  a single time and situation to another, and uses general first-order models, rather
  than guarded string models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and
  articles in \citet{ditmarsch2015handbook}.  Previous literature is motivated by epistemic
  logic and AI planning, rather computable possible worlds models in natural language semantics.
} This needs to be
implemented in a way that takes account of pre- and post-conditions for events. For
this, our approach is to refer the definition of well-formed guarded strings.
\ex{raise1} defines a relation on worlds from a relation on bare events.

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over events P and primitive
  tests B, and $\hat{R}$ be a relation on $P$. The corresponding
  relation ${R}$ on $W$ holds between a guarded string
\(
\alpha_0e_0 ... e_n\alpha_{n+1} 
\)
in $W$ and a guarded string  $q$ iff $q$ is an element of $W$ and is of the form
\(
\alpha_0'e_0' ... e_n'\alpha_{n+1}', 
\), where for $0 \leq n$, $\element{\tuple{e_i,e'_i}}{\hat{R}}$.
\end{example}

This requires that in an alternative world, each constituent event $e'_i$ is an
alternative to the event $e_i$ in the base world.
Compatibilities between events in the alternative world are enforced by the
requirement that the alternative world is an element of $W$, so that state and effect
formulas are enforced.

@@

@@

\section{The logical language of Epistemic KAT}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for KAT, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  The intended interpretation of $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet *$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet \! * \! h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet\!*\!h$ is the set
of worlds where the coin is heads.\footnote{A mapping from English sentences to logical
  terms in epistemic KAT is presented in Section 6}.

@@

A term $p$ of the logical language is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$, where superscript captures dependence
on an Epik specification.  Figure \ref{fig:interp2} defines the
interpretation.  The interpretation $\sem{ 1
}^{\epikspec}$ of the multiplicative identity 1 is the set of atoms that
satisfy the state constraint $\varphi$.  Where $b$ is a primitive
Boolean, $\sem{ b }^{\epikspec}$ is the set of atoms that satisfy the
state constraint and where $b$ is true.  Where $e$ is a primitive
event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have
the form of $e$ flanked by compatible atoms, as determined by the
event formula $\zeta_e$. The product $pq$ is
interpreted with fusion product raised to sets of guarded strings.
Kleene star is interpreted as the union of exponents ($p^n$ is
the $n$-times product of $p$ with itself, with $p^0 = 1$).
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of the embedded
proposition under the world-alternative relation ${R}_a$.

@@

Summing up, given an Epik specification $\epikspec$, term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a
set of guarded strings $\sem{p}^{\epikspec}$.  Let $K^{\epikspec}$ be
the sets that are interpretations of terms.  Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of
epistemic KAT, with operations as in Figure \ref{fig:interp2}
(e.g. the binary operation $+$ is union, and the unary operation
$\Diamond_a$ is pre-image relative to $\hat{R}_a$).  This provides a
concrete $n$-agent Kripke frame $\tuple{\sem{\bullet *}^{\epikspec},
  \hat{R}_1,...,\hat{R}_n}$.\footnote{ The domain of the Kripke frame
  differs from the domain of the guarded string model, because the
  former is the set of worlds, while the latter is the set of
  propositions.  } The frame consists of a set of worlds, and an
epistemic-alternative relation for each agent.  It is used as a target
for natural-language interpretation in Section 6.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of worlds
from the example.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
multi-character symbols that are found in implementations of the finite state
calculus. The multi-character symbol {\tt a0} is an element of the alphabet that has
no connection with the element {\tt a}.

@@

 \hspace{-13mm} \begin{tabular}[t]{cl}
{\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
$w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 1 0} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

Terms in the finite state calculus are interpreted as sets of strings,
or for relational terms, as relations between strings.
Computationally, the sets and relations are represented by finite
state acceptors.  As used here, a program in the Fst language of the
finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and
functions (defined as macros) mapping one or more regular sets or
relations to a regular set or relation.  The finite state calculus has
a product operation of string concatenation raised to sets.
Concatenation of strings with atoms (Boolean vectors) at both ends has
the effect of doubling atoms at the juncture, and does not enforce
matching of atoms at the juncture.  Therefore KAT product can not be
identified with product in the finite state calculus. Instead, KAT
product and KAT Kleene star are defined operations, see Figure
\ref{fstfigure1}.  The binary product operation {\tt Cn} and the unary
Kleene star operation {\tt Kst} combine strings in the string algebra,
remove strings with non-matching atoms, and then delete the second of
two tests to create well-formed guarded strings.  Matching of atoms is
enforced with the set {\tt Wf0}, which is the set of strings that do
not contain non-matching Boolean vectors.  The containment operator
(expressed by the dollar sign) and complement (expressed by tilde) are
operators of the finite state calculus.  The set of non-matching
sequences of atoms {\tt UnequalStPair} is defined by a finite
disjunction.  Deletion is accomplished by a re-write rule in the
finite state calculus, which is a notation for defining regular
relations by contextually constrained substitutions.  In this case, is
{\tt Squash} us a regular relation that deletes an atom (a sequence of
0's and 1's of a certain length) when it follows an atom.\footnote{
  This is a non-equal length regular relation. The finite state
  calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted
  in that the complement and set difference for non-equal length
  relations is not defined.  The epistemic alterative relations that
  are defined in Figure -- are equal-length relations.  }

@@

An event symbol as $a_1$ (Aly looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.
